---
title: "Multiple Infection Compartments"
author: "Claire P. Smith"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../R/hestia_functions.R")
source("../R/simulation.R")

library(rstan)
library(tidyverse)
library(posterior)
library(cowplot)

set.seed(6842)

```

# SIIR model (multiple infection compartments)

Lets consider an infection process model where we want to have separate compartments for symptomatic and asymptomatic infection. 

## Simulate data

First we will simulate some data from a SIIR model for 100 households. We will assume that all individuals start susceptible. The per day intra-household infection probability is 5% and the daily extra-household infection probability is 1%.

```{r}

# Simulate data from a SIIR model for 100 households
# Separate intra-household infection probabilities for each infectious compartment
# Three observations, first informative about infection (e.g. culture or PCR, 
# second informative about recovery (e.g. IgG), third informative about symptoms
dat_sim <- sim_siir(eh_prob = 0.01, ih_prob = c(0.05, 0.025), n_hh = 100,
                          hh_size = 1:5, tmax = 100, gamma = c(1/5, 1/3), split = c(0.7, 0.3),
                          covs_eh = c(0, 0), covs_ih = c(0, 0),
                          obs_prob = list(c(0.05, 0.95, 0.95, 0.05),
                                          c(0.01, 0.01, 0.01, 0.8),
                                          c(0.03, 1-1e-10, 0.03, 0.03)),
                          start_prob = c(1, 0, 0, 0),
                          complete_enroll = TRUE)
saveRDS(dat_sim, "dat_SIIR.rds")

# Name courcome columns
dat <- dat_sim$obs %>%
  rename(pcr = y1, igg = y2, symp = y3)

```

## Model specification

In addition to the data, the user needs to specify the underlying model structure, which involves two components:

1.  **The infection process model** which specifies the underlying compartmental model structure.

2.  **The observation process model** which specifies the probability of observing a positive outcome given an individual is in each compartment of the infection process model.

### Infection process model

In this infection process model, infected individuals are split into two compartments, symptomatic infections (`Is`) and asymptomatic infections (`Ia`). The proportion which goes into the symptomatic compartment is governed by the parameter `split`. Note that although there are two infectious compartments we only need to provide one value for split as the model enforces that the proportion going to the second infectious compartment is `1-split`. Here we are providing a parameter name to this argument (`split = "phi"`) indicating that we want the model to fit this parameter. We could also provide a numeric value for the model to use directly. We also use the `mult_inf_probs = TRUE` argument to indicate that we would like the two infectious compartments to have separate (rather than a single shared) intra-household infection probability.   

```{r}

# Basic SIR, fit recovery rate
inf_process <- make_infection_model(transmit(from = "S", to = c("Is", "Ia"), source = c("Is", "Ia"), split = "phi"),
                                    progress(from = "Is", to = "R", gamma_s = NA),
                                    progress(from = "Ia", to = "R", gamma_a = NA),
                                    mult_inf_probs = TRUE)


```

### Observation process model

The observation model is composed of a series of named vectors. Each vector corresponds to an observation type. Each entry in the vector is named for the corresponding compartment in the infection process model. The value is the probability of observing a positive observation given the individual is in the compartment.

In this example we have three observation types, the first which is likely to be positive when a person is actively infectious (e.g. PCR test results), the second which is more likely to be positive when the person is recovered and immune to infection (e.g. IgG antibody), and the third which is an indivator of whether someone is experiencing symptoms.

```{r}
obs_process <- make_observation_model(pcr = c("S" = 0.05, "Is" = 0.95, "Ia" = 0.95, "R" = 0.05),
                                      igg = c("S" = 0.01, "Is" = 0.01, "Ia" = 0.01, "R" = 0.8),
                                      symp = c("S" = 0.03, "Is" = 1-1e-10, "Ia" = 0.03, "R" = 0.03))

```

Now we are ready to run the model!

```{r, message=FALSE, warning=FALSE}
## Inputs
# 1. Infection process model
# 2. Observation process model
# 3. Raw data
# 4. Covariates (optional)
# 5. Starting state probabilities

res_stan <- run_model(inf_model = inf_process, obs_model = obs_process,
                      data = dat, init_probs = c(1-3*1e-10, 1e-10, 1e-10, 1e-10),
                      iter = 1000, cores = 4, save_chains = TRUE,
                      file = "../inst/stan/hmm.stan",
                      save_states = FALSE) # setting to FALSE to increase speed

saveRDS(res_stan, "SIIR_stanfit.rds")

```

Now let's look at the model results using the `posterior` package.

```{r}
# Convert to draws array
draws_full <- as_draws_array(res_stan)

# Transform variables and subset to variables of interest
draws <- subset_draws(draws_full,
                      variable = c("ih_prob[1]", "ih_prob[2]", "eh_prob", "params[1]", "params[2]", "mult_params[1]"))
draws <- mutate_variables(draws,
                          ih_prob_sym = `ih_prob[1]`,
                          ih_prob_asym = `ih_prob[2]`,
                          gamma_sym = `params[1]`,
                          gamma_asym = `params[2]`,
                          phi = `mult_params[1]`)
draws <- subset_draws(draws,
                      variable = c("ih_prob_sym", "ih_prob_asym", "eh_prob", "gamma_sym", "gamma_asym", "phi"))

# Get summary statistics
draws_sum <- summarise_draws(draws, "mean", "median",
                             ~quantile2(.x, probs = c(0.025, 0.975))) %>%
  mutate(var_type = c(rep("Infection probability", 3), rep("Recovery rate", 2), "Symptomatic proportion"))
draws_sum

# Plot versus true values
true_vals <- data.frame(var_type = c(rep("Infection probability", 3), rep("Recovery rate", 2), "Symptomatic proportion"),
                        variable = c("ih_prob_sym", "ih_prob_asym", "eh_prob", "gamma_sym", "gamma_asym", "phi"),
                        true_val = c(0.05, 0.025, 0.01, 1/5, 1/3, 0.7))

ggplot(draws_sum, aes(x = variable)) +
  geom_point(aes(y = median)) +
  geom_errorbar(aes(ymin = q2.5, ymax = q97.5)) +
  geom_hline(data = true_vals, aes(yintercept = true_val),
             color = "red", lty = "dashed") +
  facet_wrap(~var_type, scales = "free") +
  theme_bw()

```



