---
title: "Parallel and threaded fitting (including HPC)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallel and threaded fitting (including HPC)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Fitting a `hestia` model is the expensive part of most analyses. This article
explains how `hestia` uses your available cores, and how to run it efficiently
on a laptop and on a scheduled HPC cluster.

## Two levels of parallelism

A `hestia` fit can use cores in two ways at once:

1. **Across chains.** The MCMC chains are independent, so they run in parallel
   with near-linear speedup and no coordination overhead.
2. **Within a chain (threading).** The likelihood sums over households, which are
   conditionally independent, so each chain can split that sum across threads
   using Stan's `reduce_sum`.

Because chain-parallelism is "free" (no threading overhead) while within-chain
threading scales sub-linearly, `hestia` **fills chain-parallelism first and
gives any leftover cores to threads**. With `threading = TRUE` (the default)
this split is chosen for you from the cores available:

| chains | cores | chains in parallel | threads per chain |
|:------:|:-----:|:------------------:|:-----------------:|
|   4    |   4   |         4          |     1 (no spare)  |
|   4    |   8   |         4          |         2         |
|   4    |  16   |         4          |         4         |
|   2    |   8   |         2          |         4         |

So threading only "switches on" when you have more cores than chains — otherwise
every core is spent running chains.

```{r, eval=FALSE}
# Default: threading = TRUE, cores detected automatically
fit <- run_model(
  inf_model  = inf_process,
  obs_model  = obs_process,
  data       = sir,
  init_probs = c(1 - 2 * 1e-10, 1e-10, 1e-10),
  stan_opts  = stan_options(iter = 1000)
)
```

## How `hestia` decides how many cores to use

When you don't specify cores, `hestia` asks
[`parallelly::availableCores()`](https://parallelly.futureverse.org/) for the
number of cores the *process is allowed to use*. Crucially, this is **not**
`parallel::detectCores()`: it respects

- your HPC scheduler's allocation (`SLURM_CPUS_PER_TASK`, PBS, SGE, LSF),
- cgroup CPU quotas (containers, `cpuset`),
- `options(mc.cores = ...)`,

and returns 2 under `R CMD check`. This is what makes the automatic split safe
on a shared cluster (see below).

You can always take control explicitly — an explicit core count is used exactly
as given.

## On a laptop or workstation

Nothing to do: the default detects your cores and splits them. To cap usage (for
example, to keep some cores free), set `mc.cores`:

```{r, eval=FALSE}
options(mc.cores = 6)
fit <- run_model(..., stan_opts = stan_options(chains = 4))
# -> 4 chains in parallel, 1 thread each (6 %/% 4 = 1); 2 cores left free
```

## On an HPC cluster

### The trap to avoid

`parallel::detectCores()` reports **every core on the compute node** — often 64,
128, or more — regardless of how many your job was actually allocated. If a fit
sized itself from that, it would launch dozens of threads inside a job pinned to
a handful of cores, and the scheduler's cgroup would throttle it to a crawl (or
you would trample other users sharing the node). `hestia` avoids this by using
`parallelly::availableCores()`, which reads your *allocation*.

### Single node: let `hestia` split your allocation

Request one task with several CPUs, and let the automatic split divide them
between chains and threads. For 4 chains with 2 threads each (8 cores):

```bash
#!/bin/bash
#SBATCH --job-name=hestia-fit
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=8      # availableCores() will report 8
#SBATCH --mem=16G
#SBATCH --time=04:00:00

Rscript fit.R
```

```r
# fit.R
library(hestia)

fit <- run_model(
  inf_model  = inf_process,
  obs_model  = obs_process,
  data       = my_data,
  init_probs = init_probs,
  stan_opts  = stan_options(chains = 4, iter = 2000)
)
saveRDS(fit, "fit.rds")
```

No core arithmetic in your script: `availableCores()` picks up
`SLURM_CPUS_PER_TASK = 8`, and `hestia` runs 4 chains × 2 threads.

### Many chains across nodes: one chain per array task

Chain-parallelism and threads are **shared-memory**: they cannot span nodes. To
scale beyond a single node, run **one chain per job** with a SLURM array, give
each job all its cores for threading, then combine the single-chain fits
afterwards.

```bash
#SBATCH --array=1-4            # one task per chain
#SBATCH --nodes=1
#SBATCH --cpus-per-task=8      # all 8 cores -> threads for this one chain
```

```r
# each array task fits a single chain with a distinct seed
fit <- run_model(
  ...,
  stan_opts = stan_options(chains = 1, seed = Sys.getenv("SLURM_ARRAY_TASK_ID"))
)
saveRDS(fit, sprintf("chain-%s.rds", Sys.getenv("SLURM_ARRAY_TASK_ID")))
```

## Choosing the thread count yourself

`threading` accepts three forms:

- `TRUE` (default) — auto-split the available cores between chains and threads.
- `FALSE` — one thread per chain (chains still run in parallel). Useful for
  debugging or exact reproducibility comparisons.
- a **positive integer** — that many threads per chain, explicitly.

```{r, eval=FALSE}
# Exactly 2 threads per chain, whatever the machine reports
fit <- run_model(..., threading = 2)

# Single-threaded
fit <- run_model(..., threading = FALSE)
```

Parallelism lives on `run_model()`, not in `stan_options()`: `cores`,
`parallel_chains`, and `threads_per_chain` are rejected there, because
`run_model()` owns the split and translates it to whichever backend is in use.
`stan_options()` is for how the sampler explores (`iter`, `warmup`, `seed`,
`control = list(adapt_delta = ...)`, ...). Note that `threading = n` on a machine
with fewer than `n` free cores per chain will oversubscribe — that is your
explicit choice, and `hestia` honours it rather than clamping.
